---
title: "Disturbance simulations"
author: "Beni Stocker"
date: "2022-07-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(patchwork)

## install the rsofun branch 'disturb'
#remotes::install_github(repo = "computationales/rsofun", ref = "disturb")
library(rsofun)
```

## Setup

All parameters, site info, and forcing data is available for an example in the driver object (part of rsofun package)
```{r}
  biomee_p_model_drivers$forcing[[1]]
```

This contains the forcing time series data frame where the disturbance is to be defined as the fraction of aboveground biomass harvested (`harv`). Additional specifications of the disturbance forcing, which are more specific to the simulations done here, are hard-coded (see below). 


## Specify forcing

### Disturbance regime

Model is first run to steady state. Then for another 100 years undisturbed (continued steady-state) until first disturbance. After first disturbance left undisturbed for 900 years to allow for recovery (in some simulations recovery may take long). Then a regime of repeated disturbance after simulation year 1000 to investigate disturbance-recovery (non-steady state) dynamics, first at low frequency for 1000 years (disturbance every 250 years), then at high frequency for 1000 years (disturbance every 25 years).

Disturbance is implemented by:
- year 100 first disturbance applied, then recovery until year 1000
- year 1000 second disturbance, then every 250 years disturbed for 1000 years
- ... then every 25 years disturbed for another 1000 years

To be handled by model by (new) forcing time series as input (`harv`)
```{r}
fharv <- 0.9
harv_vec <- rep(0, 999)
harv_vec[100] <- fharv
harv_vec <- c(harv_vec, rep(c(fharv, rep(0, 249)), 4), rep(c(fharv, rep(0, 24)), 40), 0)

df_harv <- tibble(year = seq(length(harv_vec)), harv = harv_vec)

df_harv %>% 
  ggplot(aes(year, harv)) +
  geom_line() +
  ylim(0, 1)
```

### Experimental design

```{r}
df_sims <- read.csv("~/disturb/simulations_disturb.csv")
df_sims %>% knitr::kable()
```

### Create forcing objects

```{r}
## get mean seasonal cycle and repeat this every year of all simulations
df_forcing <- biomee_p_model_drivers$forcing[[1]] %>% 
  mutate(doy = lubridate::yday(date)) %>% 
  group_by(doy) %>% 
  summarise(across(is.numeric, mean))
```

Repeat mean seasonal cycle `nyears` times where `nyears` corresponds to the length of the harvest time series (rows in `df_harv`). The column `year` now signifies simulation year and goes from 1 to `nyears`.

Add harvest forcing to drivers. 
```{r}
nyears <- nrow(df_harv)

df_forcing <- df_forcing %>% 
  slice(rep(1:n(), nyears)) %>% 
  mutate(year = rep(1:nyears, each = 365))
```

Add harvest to forcing, assuming harvest on Jan 1st.
```{r}
df_forcing_disturb <- df_forcing %>% 
  left_join(
    df_harv %>% 
      mutate(doy = 1),
    by = c("doy", "year")
  ) %>% 
  mutate(harv = ifelse(is.na(harv), 0, harv))
  
  # ## add pseudo-date, starting in year 2000
  # mutate(date = lubridate::ymd("0000-01-01") + lubridate::years(year-1) + lubridate::days(doy-1))

## for control simulation
df_forcing <- df_forcing %>%
  mutate(harv = 0)
```

Add N deposition as NOx and NHy.
```{r}
df_forcing <- df_forcing %>% 
  mutate(nox = 0, nhy = 0)

df_forcing_disturb <- df_forcing_disturb %>% 
  mutate(nox = 0, nhy = 0)
```

Create new driver objects.
```{r}
## control simulations without disturbance
biomee_p_model_drivers_xx0 <- biomee_p_model_drivers
biomee_p_model_drivers_xx0$forcing[[1]] <- df_forcing
biomee_p_model_drivers_xx0$params_siml[[1]]$firstyeartrend <- 0
biomee_p_model_drivers_xx0$params_siml[[1]]$nyeartrend <- 3000

## simulations with disturbance
biomee_p_model_drivers_xx1 <- biomee_p_model_drivers
biomee_p_model_drivers_xx1$forcing[[1]] <- df_forcing_disturb
biomee_p_model_drivers_xx1$params_siml[[1]]$firstyeartrend <- 0
biomee_p_model_drivers_xx1$params_siml[[1]]$nyeartrend <- 3000
```


## Model runs

### sc1

```{r}
filn <- "data/out_sc1.rds"
if (!file.exists(filn)){
  out_sc1 <- runread_biomee_f(
       biomee_p_model_drivers_xx0,
       makecheck = TRUE,
       parallel = FALSE
       )
  saveRDS(out_sc1, file = filn)
} else 
  out_sc1 <- readRDS(filn)
```

### sc2

Export of dead biomass from system (not added to soil) is implemented by simply not calling the `plant2soil()` in the subroutine `disturb()` (file `vegetation_biomee.mod.f90`). Comment it out and re-compile before running.
```{r}
filn <- "data/out_sc2.rds"
if (!file.exists(filn)){
  out_sc2 <- runread_biomee_f(
       biomee_p_model_drivers_xx1,
       makecheck = TRUE,
       parallel = FALSE
       )
  saveRDS(out_sc2, filn)
} else 
  out_sc2 <- readRDS(filn)
```

## Analysis

Take annual tile-level outputs (for shorthand).
```{r}
out_sc1_ann <- out_sc1$data[[1]]$output_annual_tile
out_sc2_ann <- out_sc2$data[[1]]$output_annual_tile
```


### Plant C

```{r}
# model output includes the spinup. Remove it for plotting and overwrite years.
out_sc1_ann <- out_sc1_ann %>%
  slice((biomee_p_model_drivers_xx1$params_siml[[1]]$spinupyears + 1):nrow(out_sc1_ann)) %>% 
  mutate(year = 1:nyears)
out_sc2_ann <- out_sc2_ann %>%
  slice((biomee_p_model_drivers_xx1$params_siml[[1]]$spinupyears + 1):nrow(out_sc2_ann)) %>% 
  mutate(year = 1:nyears)
  
gg_sc1 <- out_sc1_ann %>% 
  ggplot() +
  geom_line(aes(x = year, y = plantC)) +
  theme_classic() +
  geom_vline(xintercept = df_harv %>% filter(harv > 0) %>% pull(year), color = "red", alpha = 0.3) +
  labs(x = "Year", y = "plant C", title = "Simulation sc1") +
  ylim(0, 15)

gg_sc2 <- out_sc2_ann %>% 
  ggplot() +
  geom_line(aes(x = year, y = plantC)) +
  theme_classic() +
  geom_vline(xintercept = df_harv %>% filter(harv > 0) %>% pull(year), color = "red", alpha = 0.3) +
  labs(x = "Year", y = "plant C", title = "Simulation sc2") +
  ylim(0, 15)

gg_sc1 /
  gg_sc2
```

### Soil C

```{r}
gg_sc1 <- out_sc1_ann %>% 
  ggplot() +
  geom_line(aes(x = year, y = fastSOM + SlowSOM)) +
  theme_classic() +
  geom_vline(xintercept = df_harv %>% filter(harv > 0) %>% pull(year), color = "red", alpha = 0.3) +
  labs(x = "Year", y = "Soil C", title = "Simulation sc1") +
  ylim(0, 100)

gg_sc2 <- out_sc2_ann %>% 
  ggplot() +
  geom_line(aes(x = year, y = fastSOM + SlowSOM)) +
  theme_classic() +
  geom_vline(xintercept = df_harv %>% filter(harv > 0) %>% pull(year), color = "red", alpha = 0.3) +
  labs(x = "Year", y = "Soil C", title = "Simulation sc2") +
  ylim(0, 100)

gg_sc1 /
  gg_sc2
```
